---
title: "Terrorism in the US, 1994-2014"
author: "David Himrich"
date: "December 5, 2015"
output: html_document
---

# Introduction

This document is an analysis of a portion of the Global Terrorism Database (GTD) maintained by the National Consortium for the Study of Terrorism and Responses to Terrorism (START). The homepage for the database is [here](http://www.start.umd.edu/gtd/). Users may search the database from that page. Users may also download the database or portions thereof. This analysis is based on database from 1992 through 2014, which is distributed as two Excel files, gtd\_92to10\_0615dist.xlsx and gtd\_11to14\_0615dist.xlsx. 

## About this Document

The HTML file for this analysis is created in RStudio using [R Markdown](rmarkdown.rstudio.com) and the knitr package for the R statistical programming language. The R commands to produce the analysis are embedded in the document. RStudio users will be able to reproduce this study by copying the two .csv files exported from the Excel files described above as well as the US\_terror.Rmd file into the working directory and using the knitr package to produce an HTML output document from the .Rmd file. The Github repository for this project is [https://github.com/dhimrich/Terrorism.git]. 


# Data Preparation


```{r data.prep, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE}
    


## Terrorism Data Prep
gtd_92_10.df <- read.csv("gtd_92to10_0615dist.csv")

library(dplyr)

us_92_10.df <- gtd_92_10.df %>% 
    filter(country_txt == "United States")

gtd_11_14.df <- read.csv("gtd_11to14_0615dist.csv")

us_11_14.df <- gtd_11_14.df %>% 
    filter(country_txt == "United States")

us_94_14.df <- bind_rows(us_92_10.df, us_11_14.df)
us_94_14.df <- filter(us_94_14.df, iyear >= 1994)


rm(gtd_11_14.df, gtd_92_10.df, us_11_14.df, us_92_10.df)


## Create a variable for perpetrator categories and assign unknowns
us_94_14.df$perp.cat <- NA

## Assign some perp.cat values based on target type
us_94_14.df$perp.cat[us_94_14.df$targtype1_txt %in% 
                         c("Abortion Related")] <- "anti-abortion"

## Assign other perp.cat values based on group name (gname)
us_94_14.df$perp.cat[us_94_14.df$gname %in% 
                         c("Anti-Abortion Activists", 
                           "Army of God")] <- "anti-abortion"

us_94_14.df$perp.cat[us_94_14.df$gname %in% 
                         c("Earth Liberation Front (ELF)",
                           "Animal Liberation Front (ALF)",
                           "The Justice Department",
                           "Coalition to Save the Preserves (CSP)",
                           "Revolutionary Cells-Animal Liberation Brigade",
                           "Farm Animal Revenge Militia (FARM)",
                           "Earth First!",
                           "Animal Rights Activists",
                           "Anarchists")] <- "eco-terror"

us_94_14.df$perp.cat[us_94_14.df$gname %in% 
                         c("Aryan Republican Army",
                           "Maccabee Squad and the Shield of David",
                           "Jewish Extremists",
                           "Cuban Exiles",
                           "Sons of the Gestapo")] <- "right-wing"

us_94_14.df$perp.cat[us_94_14.df$gname %in% 
                         c("World Church of the Creator",
                           "Ku Klux Klan",
                           "Fourth Reich Skinheads")] <- "anti-minority"

us_94_14.df$perp.cat[us_94_14.df$gname %in% 
                         c("Al-Qa`ida",
                           "Tehrik-i-Taliban Pakistan (TTP)",
                           "Islamist Extremists",
                           "Al-Qa`ida in the Arabian Peninsula (AQAP)")] <- "jihadis"

us_94_14.df$perp.cat[us_94_14.df$gname %in% 
                         c("Veterans United for Non-Religious Memorials")] <- "anti-Christian"

us_94_14.df$perp.cat[us_94_14.df$gname %in% 
                         c("Mexican Revolutionary Movement")] <- "left-wing"

us_94_14.df$perp.cat[us_94_14.df$gname %in% 
                         c("Aryan Nation")] <- "anti-Jew"





## Assign individual and unknown groups based on reading summaries of individual
## incidents

us_94_14.df$perp.cat[us_94_14.df$eventid %in% 
                         c("199410150011", "199504150010",
                          "199608160007", "199905000002",
                          "200003250005", "200005270009",
                          "200006160006", "200105050003",
                          "200305040005", "200309240007",
                          "200310250002", "200412090005",
                          "200501010007", "200502130002",
                          "200502150001", "200502170002",
                          "200507080006", "201009010022",
                          "201110120003")] <- "eco-terror"

us_94_14.df$perp.cat[us_94_14.df$eventid %in% 
                         c("200202040010", "200603030013",
                          "200906010028", "200911060002",
                          "201010000001", "201010000002",
                          "201010190014", "201010250026",
                          "201011000001", "201304150001",
                          "201304150002", "201304180001",
                          "201304190009", "201404270057",
                          "201406010063", "201406250082",
                          "201410230047")] <- "jihadis"

us_94_14.df$perp.cat[us_94_14.df$eventid %in% 
                         c("199601120006")] <- "anti-Christian"

us_94_14.df$perp.cat[us_94_14.df$eventid %in% 
                         c("199906180002", "199906180003",
                          "199906180004", "200603030013",
                          "200804220011", "200906100003",
                          "201404130060", "201404130061",
                          "200311180007", "200404010004")] <- "anti-Jew"

us_94_14.df$perp.cat[us_94_14.df$eventid %in% 
                         c("200512200004", "201005100042", 
                          "201005100042", "201104230010",
                          "201201030020", "201201030019",
                          "201201030021", "201201030022",
                          "201207040032", "201208060019",
                          "201208120012", "201209300041",
                          "201403250090", "200802090004")] <- "anti-Muslim"

us_94_14.df$perp.cat[us_94_14.df$eventid %in% 
                         c("199601120006", "200811050008", 
                          "200907030004", "201101170018",
                          "201208050006", "199610030012",
                          "199907010009", "199410160011",
                          "199410160012", "199410160013",
                          "200811140015")] <- "anti-minority"


us_94_14.df$perp.cat[us_94_14.df$eventid %in% 
                         c("199804140001", "200609110007",
                           "199907020007")] <- "anti-abortion"

us_94_14.df$perp.cat[us_94_14.df$eventid %in% 
                         c("199903200009", "200210030002",
                          "200807250030", "200807270001",
                          "201206180029", "201311010046",
                          "201409120032", "201411280018",
                          "201411280019", "201411280020",
                          "199504190004")] <- "right-wing"

us_94_14.df$perp.cat[us_94_14.df$eventid == "199504190004"] <- "OKC"

us_94_14.df$perp.cat[us_94_14.df$eventid %in% 
                         c("200206030002", "200206030003",
                          "201002180013", "201208150059",
                          "201302030025", "201304110003", 
                          "201406080071", "201410030065",
                          "201409110001", "201412200060")] <- "left-wing"

unks <- us_94_14.df %>% 
    filter(gname %in% c("Individual", "Unknown")) %>% 
    select(eventid, targtype1_txt, summary, addnotes)


```

I produced the data set for this analysis from the two files listed above by filtering them to only records for the United States, and then combining the remaining records. The resulting file has `r nrow(us_94_14.df)` observations of `r ncol(us_94_14.df)` variables. 

I wanted to study the number and timing of incidents in general categories such as jihadis, anti-abortion terrorists, racially-oriented terrorists, and so forth. With the exception of anti-abortion incidents, which are identified by a target type designated "Abortion Related", the database as distributed does not identify perpetrator categories of this type. The *gname* variable contains a group name for some incidents, but in this US data set, `r nrow(unks)` incidents are either "Unknown" or "Individual".

I proceeded by creating the *perp.cat* variable. I assigned incidents to categories of my choosing based on reading the *summary* and *motive* text fields, assisted by internet searches as necessary. I ended up with the following categories, shown by descending number of incidents, along with the number of killed, wounded, and the number of total casualties (killed plus wounded.)


```{r perp.display, echo=FALSE, warning=FALSE, message=FALSE}

library(dplyr)
us_94_14.df %>% 
    group_by(perp.cat) %>% 
    summarize(incidents = NROW(eventid),
              casualties = 
                  sum(nkill, na.rm = TRUE) + 
                  sum(nwound, na.rm = TRUE),
              killed = sum(nkill, na.rm = TRUE),
              wounded = sum(nwound, na.rm = TRUE)) %>% 
    arrange(desc(incidents)) %>% knitr::kable()

```


Unfortunately there remain a large number of incidents with insufficient information in the database to allow them to be reliably categorized. We could probably guess fairly accurately at categories from some of these group names, but since those incident and casualty counts are quite small, I chose to exclude these from further analysis.

I included the 1995 bombing of the Murrah Building in Oklahoma City in its own "OKC" category. That incident produced the fourth-largest number killed in this database, trailng only the destruction of the two World Trade Center towers and the attack on the Pentagon on 9/11/01. In the analysis section I will present the data in way that I believe allows the reader to compare or exclude these large events in ways that are meaningful from varying perspectives.



```{r na.table, echo=FALSE, warning=FALSE, message=FALSE}

us_94_14.df %>%
    filter(is.na(perp.cat)) %>% 
    group_by(gname) %>% 
    summarize(incidents = NROW(eventid),
              casualties = 
                  sum(nkill, na.rm = TRUE) + 
                  sum(nwound, na.rm = TRUE),
              killed = sum(nkill, na.rm = TRUE),
              wounded = sum(nwound, na.rm = TRUE)) %>% 
    arrange(desc(incidents)) %>% knitr::kable()


```

# Analysis

My main questions of interest are how these categories of terror incidents compare to each other in terms of number of incidents and casualties, and how they have varied over time. I will present these comparisons via some simple graphics.

## Perpetrator Category Comparison

```{r inc.bar, echo=FALSE, warning=FALSE, message=FALSE}

library(ggplot2)

## Make a data frame just for plotting
plot.df <- us_94_14.df %>% 
    filter(!is.na(perp.cat)) %>% 
    select(eventid, iyear, imonth, iday, nkill, nwound, perp.cat) 

## There are some NA values in nwound, set those to zero
plot.df$nwound[is.na(plot.df$nwound)] <- 0
    
## Sum the killed and wounded to get casualties
plot.df <- mutate(plot.df, casualties = nkill + nwound)

bi <- ggplot(plot.df, aes(perp.cat))
bi + geom_bar() + ggtitle("Incidents by Perpetrator Category") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 0.8)) +
    xlab("Perpetrator Category") + ylab("Incident Count")

```

We see that in terms of incident counts, the "eco-terror" category, in which I have included both environmentalist and animal-rights terrorist incidents, and the "anti-abortion" category predominate.

In terms of risk perception, we are probably more interested in the number of casualties:

```{r cas.bar, echo=FALSE, warning=FALSE, message=FALSE}

bc <- ggplot(plot.df, aes(perp.cat, casualties))
bc + geom_bar(stat = "identity") + 
    ggtitle("Casualties by Perpetrator Category") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 0.8)) +
    xlab("Perpetrator Category") + ylab("Casualties")

```

As expected, the casualties due to jihadi actions dwarf the other categories, mostly due to the attacks of September 11, 2001, which are recorded in this database as four incidents. Also, the total number wounded is not recorded for the destruction of the World Trade Center, which certainly is a signficant distortion to the casualy figures. We can exclude those four incidents and compare the remaining fatality totals:

```{r kill.bar, echo=FALSE, warning=FALSE, message=FALSE}

plot2.df <- plot.df %>% 
    filter(!(perp.cat == "jihadis" & grepl("^2001", eventid)))

bc2 <- ggplot(plot2.df, aes(perp.cat, nkill))
bc2 + geom_bar(stat = "identity") + 
    ggtitle("Fatalities by Perpetrator Category (excl. 9/11/01)") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 0.8)) +
    xlab("Perpetrator Category") + ylab("Fatalities")

plot2.df %>%
    group_by(perp.cat) %>% 
    summarize(Fatalities = sum(nkill)) %>% 
    arrange(desc(Fatalities)) %>% knitr::kable()

```

Having set aside the most consequential terrorist event in US history (the 9/11/01 attacks), we can gain better insight into what remains. We see that our other singular event, the Oklahoma City bombing in April 1995 far overshadows the other categories in terms of fatalities. The much talked-about backlash against Muslims in the US has mostly failed to materialize: I have been able to identify 13 such incidents in this database between 1992 and 2014. Those incidents have produced no casualties.

## Incidents Over Time

The number of incidents in most of the categories seems limited. We might be more interested in the number of these incidents over time. Let's look at the incidents counts by year for some of the larger categories

```{r i.year1, echo=FALSE, warning=FALSE, message=FALSE}
## Make a data frame just for plotting
## Set up a column of all the years and all the perp.cat values
plot3.df <- data.frame(iyear = rep(1994:2014, 
                                   length(unique(us_94_14.df$perp.cat))))
plot3.df <- arrange(plot3.df, iyear)
plot3.df$perp.cat <- unique(us_94_14.df$perp.cat)

plot3a.df <- us_94_14.df %>% 
    group_by(perp.cat, iyear) %>% 
    summarize(Incidents = NROW(eventid))

plot3b.df <- left_join(plot3.df, plot3a.df)

## Fix some NA values for plotting
plot3b.df$perp.cat[is.na(plot3b.df$perp.cat)] <- "N/A"
plot3b.df$Incidents[is.na(plot3b.df$Incidents)] <- 0

yi1 <- ggplot(filter(plot3b.df, perp.cat %in% 
                        c("eco-terror", "anti-abortion", "N/A")), 
                    aes(iyear, Incidents, color = perp.cat))

yi1 + geom_line(aes(group=perp.cat), size=2) + 
    scale_color_brewer(type = "qual", palette = "Set1") +
    ggtitle("Incident Count by Year") + 
    xlab("Year of Incident")


```

All of these large categories appear to be less frequent in recent years than in earlier years. Let's look at a similar chart of the next largest categories:

```{r i.year2, echo=FALSE, warning=FALSE, message=FALSE}
## Make a data frame just for plotting

yi2 <- ggplot(filter(plot3b.df, perp.cat %in% 
                        c("right-wing", "jihadis", 
                          "anti-minority", "anti-Muslim")), 
                    aes(iyear, Incidents, color = perp.cat))

yi2 + geom_line(aes(group=perp.cat), size=2) + 
    scale_color_brewer(type = "qual", palette = "Set1") +
    ggtitle("Incident Count by Year") + 
    xlab("Year of Incident")

library(tidyr)
plot3b.df %>% 
    filter(perp.cat %in% c("right-wing", "jihadis", 
                          "anti-minority", "anti-Muslim")) %>% 
    spread(perp.cat, Incidents)%>% 
    knitr::kable()

```

Incidents in these smaller categories have occurred sporadically since 1994. It's not clear from the chart that any were increasing in frequency through 2014.

# Conclusion

From 1994 Through 2014, the incidence of terrroris acts in the US appears to have decreased. This seems to be consistent with the general trend for violent crime in the US during this period. There seems to be no evidence of a widespread backlash against Muslims.

The terrorist attacks with the greatest impact are of course those producing mass casualties. These events are few in number (essentially 5 events out of 566 in this data set) and seemed to arrive "out of the blue". As I write this in December, 2015 it's not clear that the immediate future is going to be like the recent past. A string of jihadi acts that build on events since 2013 may alter the security environment in the US considerably.


